UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-07-04 ms/main.tex
edits up to line 159

Focusing on UMAP with the leading 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), several population clusters reflect shared ancestries. British individuals from England and Scotland form a cluster mixed with those from Utah who claim Northern and Western European ancestry. Toscani and Iberian individuals form a cluster reflecting their Mediterranean heritage. African Americans in the Southwest US, African Caribbean individuals in Barbados, and some Puerto Ricans also form a cluster. Three East Asian clusters appear: one is largely Han and Southern Han individuals, another is comprised of the Chinese Dai in southern China and the Kinh from Vietnam, and the third is the Japanese population. Other clusters are comprised of Colombians and Peruvians, the Esan and Yoruba populations of Nigeria, and several South Asian populations. Within population clusters, family members were projected near each other but did not form disjoint clusters unless UMAP was parameterized to use a minimal number of neighbours (figure~\ref{fig:supp_1kgp_families}).

@@

\subsubsection*{Population structure in the UKBB reflects local and global genetic variation} 
The UKBB contains data on 488,377 individuals containing phenotypic measures and self-identified ethnic backgrounds. Comparing UMAP to PCA with this data, the former captures local and fine-scale structure well while the latter identifies major axes of variation in figure~\ref{fig:fig_ukbb}. UMAP on the top 10 principal components reveals continuous and discrete population structure (figure~\ref{fig:umap_ukbb_pc10}): the patchwork of local topologies identifies continuous structure within the British population as well as admixture gradients despite the very unbalanced population sizes. The result is a comprehensive portrait of genetic variation capturing population relationships not visible using other methods, succinctly illustrating the complex structure of a large and multi-ethnic dataset.

>>>>

Focusing on UMAP with the leading 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), several population clusters reflect shared ancestries. British individuals from England and Scotland form a cluster mixed with those from Utah who claim Northern and Western European ancestry. Toscani and Iberian individuals form a cluster reflecting their Mediterranean heritage. African Americans in the Southwest US, African Caribbean individuals in Barbados, and some Puerto Ricans also form a cluster. Three East Asian clusters appear: one is largely Han and Southern Han individuals, another is comprised of the Chinese Dai in southern China and the Kinh from Vietnam, and the third is the Japanese population. Other clusters are comprised of Colombians and Peruvians, the Esan and Yoruba populations of Nigeria, and several South Asian populations.

Within population clusters, family members were projected near each other within broader population groups. When UMAP was parameterized to use only 5 nearest neighbours, however, families often formed distinct clusters (figure~\ref{fig:supp_1kgp_families}): Using few neighbors to build a manifold emphasizes closer relatedness.

@@

\subsubsection*{Population structure in the UKBB reflects local and global genetic variation} 
The UKBB contains data on 488,377 individuals containing phenotypic measures and self-identified ethnic backgrounds. Figure~\ref{fig:fig_ukbb} compares UMAP to PCA apply to UKBB. As expected, PCA captures major axes of variation emphasizing continental ancestry, whereas UMAP reveals finer sturcture. UMAP on the top 10 principal components reveals continuous and discrete population structure (figure~\ref{fig:umap_ukbb_pc10}): the patchwork of local topologies identifies multiple sub-populations, as well as continuous structure within populations and admixture gradients between populations. The result is a comprehensive portrait of genetic variation capturing population relationships not visible using other methods, succinctly illustrating the complex structure of a large and multi-ethnic dataset.
